---
title: "models"
output: html_document
---

# Connection

```{r conn, include=FALSE}
# Connect to database
library(DBI)
library(RSQLite)
library(dplyr)
library(tidyr)
library(ggplot2)
library(car)
library(glmnet)
library(caret)
library(rpart)
library(reshape2)
library(knitr)
library(randomForest)
library(gbm)
library(mgcv)
library(kableExtra)
library(webshot2)
library(flextable)
library(gdtools)
library(piecewiseSEM)
library(fastshap)
library(shapviz)
library(ranger)
library(pdp)
library(Hmisc)

setwd("C:/Users/kelto/GT_Analytics/Spring_2026/Data_Mining/ISYE7406_GITHUB_REPO")
con <- dbConnect(SQLite(), "NLA.db")

df <- dbReadTable(con, "rel_vars")
```

# Feature Engineering

```{r engineer}
## clay = 1 - sand - silt
df$calc_clay <- 1 - (df$BSFCSAND + df$BSFCSILT)
hist(df$calc_clay)
hist(df$BSFCSAND + df$BSFCSILT)

# log transform
## chlorophyll a
hist(df$CHLA_RESULT)
df$log_chla <- log10(df$CHLA_RESULT)
hist(df$log_chla)

## nitrogen
hist(df$NTL_RESULT)
df$log_N <- log10(df$NTL_RESULT)
hist(df$log_N)

## phosphorus
hist(df$PTL_RESULT)
df$log_P <- log10(df$PTL_RESULT)
hist(df$log_P)

## herb bio
hist(df$HERB_BIO)
df$log_herb_bio <- log10(df$HERB_BIO+1)
hist(df$log_herb_bio)

## omni bio
hist(df$OMNI_BIO)
df$log_omni_bio <- log10(df$OMNI_BIO+1)
hist(df$log_omni_bio)

## pred bio
hist(df$PRED_BIO)
df$log_pred_bio <- log10(df$PRED_BIO+1)
hist(df$log_pred_bio)

## benthic filterers
hist(df$COFIPIND)
df$log_ben_filt <- log10(df$COFIPIND+1)
hist(df$log_ben_filt)

## benthic gatherers
hist(df$COGAPIND)
df$log_ben_gath <- log10(df$COGAPIND+1)
hist(df$log_ben_gath)

## benthic predators
hist(df$PREDPIND)
df$log_ben_pred <- log10(df$PREDPIND+1)
hist(df$log_ben_pred)

## benthic shredders
hist(df$SHRDPIND)
df$log_ben_shrd <- log10(df$SHRDPIND+1)
hist(df$log_ben_shrd)

## benthic scrapers
hist(df$SCRPPIND)
df$log_ben_scrp <- log10(df$SCRPPIND+1)
hist(df$log_ben_scrp)

```

# Feature Selection

```{r select}
all_vars <- df %>% 
  subset(select = c(log_turb, # L0
    calc_clay, RVFPCANDECIDUOUS_RIP, log_chla, AMITOTAL, # L1
    log_N, log_P, TEMPERATURE, log_herb_bio, log_omni_bio, log_pred_bio, log_ben_filt, log_ben_gath, log_ben_pred, log_ben_shrd, log_ben_scrp, Accipitriformes_nspecies, Anseriformes_nspecies, Coraciiformes_nspecies, Pelecaniformes_nspecies # L2
    ))
```

# Nans

```{r nans}
# remove chunk of nans
all_vars <- all_vars %>% filter(!is.na(calc_clay))
boxplot(colSums(is.na(all_vars)), horizontal = TRUE, xlab="# NAs per feature", main="NAs remaining")

# omit all nas
print(paste("Nrows in dataset with all NAs omitted:", nrow(na.omit(all_vars))))

all_vars <- na.omit(all_vars)
```

# Multicollinearity
And outliers?

```{r corr}
## **AI generated figure**

# compute correlations & p-values
cor_res <- rcorr(as.matrix(all_vars))  # replace df with your data
r <- cor_res$r
p <- cor_res$P

# lower triangle only
r[upper.tri(r, diag = TRUE)] <- NA
p[upper.tri(p, diag = TRUE)] <- NA

melted <- merge(
  reshape2::melt(r, na.rm = TRUE, value.name = "r"),
  reshape2::melt(p, na.rm = TRUE, value.name = "p")
)

# significance stars
melted$label <- paste0(
  sprintf("%.2f", melted$r),
  ifelse(melted$p < .001, "***", ifelse(melted$p < .01, "**", ifelse(melted$p < .05, "*", "")))
)

p <- ggplot(melted, aes(Var2, Var1, fill = r)) +
  geom_tile(color = "white") +
  # geom_text(aes(label = label), size = 3) +
  scale_fill_gradient2(low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
                       limits = c(-1, 1), name = "r") +
  coord_fixed() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(x = NULL, y = NULL)

ggsave("figures/corrplot.png", plot = p)
```


```{r multi}
# VIF
vif_lm_turb <- lm(log_turb~.,data = all_vars %>% subset(select=c(log_chla, calc_clay, RVFPCANDECIDUOUS_RIP, log_turb)))
v <- vif(vif_lm_turb)
print(sort(v, decreasing = TRUE))

vif_lm_chla <- lm(log_chla~.,data = all_vars %>% subset(select=c(log_chla, log_N, log_P, TEMPERATURE, log_herb_bio, log_omni_bio, log_pred_bio, log_ben_filt, log_ben_gath, log_ben_pred, log_ben_shrd, log_ben_scrp, Accipitriformes_nspecies, Anseriformes_nspecies, Coraciiformes_nspecies, Pelecaniformes_nspecies)))
v <- vif(vif_lm_chla)
print(sort(v, decreasing = TRUE))

vif_lm_algae <- lm(AMITOTAL~.,data = all_vars %>% subset(select=c(AMITOTAL, log_N, log_P, TEMPERATURE, log_herb_bio, log_omni_bio, log_pred_bio, log_ben_filt, log_ben_gath, log_ben_pred, log_ben_shrd, log_ben_scrp, Accipitriformes_nspecies, Anseriformes_nspecies, Coraciiformes_nspecies, Pelecaniformes_nspecies)))
v <- vif(vif_lm_algae)
print(sort(v, decreasing = TRUE))
```

# Model Definitions

```{r formulas}
# formulas
m1 <- as.formula("log_turb~calc_clay+RVFPCANDECIDUOUS_RIP+log_chla")
m2 <- as.formula("log_chla~log_N+log_P+TEMPERATURE+log_herb_bio+log_omni_bio+log_pred_bio+log_ben_filt+log_ben_gath+log_ben_pred+log_ben_shrd+log_ben_scrp+Accipitriformes_nspecies+Anseriformes_nspecies+Coraciiformes_nspecies+Pelecaniformes_nspecies")
m3 <- as.formula("AMITOTAL~log_N+log_P+TEMPERATURE+log_herb_bio+log_omni_bio+log_pred_bio+log_ben_filt+log_ben_gath+log_ben_pred+log_ben_shrd+log_ben_scrp+Accipitriformes_nspecies+Anseriformes_nspecies+Coraciiformes_nspecies+Pelecaniformes_nspecies")
```


# Cross Validation

We want to compare confidence intervals for model performance (R2) and feature importance between the random forest and piecewise SEM models. We will do this with 100 rounds of Monte Carlo CV with p = 0.8.

```{r cv}
t0 <- Sys.time()

# setup
B <- 100
N <- nrow(all_vars)
p <- 0.8
S <- round(p*N)
rf1 <- vector("list", B)
rf2 <- vector("list", B)
rf3 <- vector("list", B)
sem0 <- vector("list", B)
set.seed(10)

# pred grid
all_vars_std <- as.data.frame(scale(all_vars))
col_max <- apply(all_vars_std, 2, max)
col_min <- apply(all_vars_std, 2, min)
col_seqs <- Map(function(mn, mx) seq(mn, mx, length.out = 20), col_min, col_max)

# functions
fit_ranger <- function(f, d) {
  return(ranger(f, data = d, importance = "permutation"))
}

pred_ranger <- function(object, newdata) {
  return(predict(object, data = newdata)$predictions)
}

calc_r2 <- function(pred, test_vals, train_vals) {
  ssres <- sum((test_vals - pred)^2)
  sstot <- sum((test_vals - mean(train_vals))^2)
  return(1 - (ssres/sstot))
}

calc_partials <- function(m, pvars, pfun, d) {
  pdps <- vector("list", length(pvars))
  for (v in seq(length(pvars))) {
    v_grid <- data.frame(col_seqs[[pvars[v]]])
    colnames(v_grid) <- pvars[v]
    pdps[[v]] <- list(
      Var = pvars[v],
      PDP = partial(m, train = d, pred.var = c(pvars[v]), pred.fun = pfun, pred.grid = v_grid, type = "regression", plot = FALSE, chull = TRUE)
    )
  }
  return(pdps)
}

# MC CV loop
for (i in seq(B)) {
  l0 <- Sys.time()
  
  # split
  mask <- sample(1:N, S)
  train <- all_vars[mask,]
  test <- all_vars[-mask,]
  
  # standardize
  train_means <- colMeans(train)
  train_sds <- apply(train, 2, sd)
  train_std <- as.data.frame(scale(train, center = train_means, scale = train_sds))
  test_std <- as.data.frame(scale(test, center = train_means, scale = train_sds))
  
  # forest
  ## L1
  rf <- fit_ranger(m1, train_std)
  pred <- pred_ranger(rf, test_std)
  r2 <- calc_r2(pred, test_std$log_turb, train_std$log_turb)
  pdps <- calc_partials(rf, labels(terms(m1)), pred_ranger, train_std)
  
  rf1[[i]] <- list(
    R2 = r2,
    imp = rf$variable.importance,
    pdp = pdps
  )
  
  ## L2
  rf <- fit_ranger(m2, train_std)
  pred <- pred_ranger(rf, test_std)
  r2 <- calc_r2(pred, test_std$log_chla, train_std$log_chla)
  pdps <- calc_partials(rf, labels(terms(m2)), pred_ranger, train_std)

  rf2[[i]] <- list(
    R2 = r2,
    imp = rf$variable.importance,
    pdp = pdps
  )

  ## L3
  rf <- fit_ranger(m3, train_std)
  pred <- pred_ranger(rf, test_std)
  r2 <- calc_r2(pred, test_std$AMITOTAL, train_std$AMITOTAL)
  pdps <- calc_partials(rf, labels(terms(m3)), pred_ranger, train_std)

  rf3[[i]] <- list(
    R2 = r2,
    imp = rf$variable.importance,
    pdp = pdps
  )
  
  # pSEM
  lm1 <- lm(m1, data = train_std)
  lm2 <- lm(m2, data = train_std)
  lm3 <- lm(m3, data = train_std)

  s0 <- psem(lm1, lm2, lm3, log_chla %~~% AMITOTAL, data = train_std)
  
  ## L1
  pred <- predict(lm1, newdata = test_std)
  r2_1 <- calc_r2(pred, test_std$log_turb, train_std$log_turb)

  ## L2
  pred <- predict(lm2, newdata = test_std)
  r2_2 <- calc_r2(pred, test_std$log_chla, train_std$log_chla)

  ## L3
  pred <- predict(lm3, newdata = test_std)
  r2_3 <- calc_r2(pred, test_std$AMITOTAL, train_std$AMITOTAL)

  sem0[[i]] <- list(
    R2 = list(R2_1 = r2_1, R2_2 = r2_2, R2_3 = r2_3),
    imp = summary(s0, .progressBar = FALSE)$coefficients
  )
  
  l1 <- Sys.time()
  print(paste("MC CV loop time:", round(l1-l0, 4), ", B:", i))

}

t1 <- Sys.time()
print(paste("Execution time:", round(t1-t0, 4), ", Total MC CV rounds:", B))
```

# Model Comparison
Perform Wilcoxon signed-rank and t-tests on R2 for each model

```{r model comp}
r2_rf <- data.frame(
  M1 = numeric(0),
  M2 = numeric(0),
  M3 = numeric(0)
)

r2_sem <- data.frame(
  M1 = numeric(0),
  M2 = numeric(0),
  M3 = numeric(0)
)

# forest
for (i in seq(B)) {
  r2_1 <- rf1[[i]]$R2
  r2_2 <- rf2[[i]]$R2
  r2_3 <- rf3[[i]]$R2
  temp <- data.frame(M1 = r2_1, M2 = r2_2, M3 = r2_3)
  r2_rf <- rbind(r2_rf, temp)
}

# psem
for (i in seq(B)) {
  r2_1 <- sem0[[i]]$R2$R2_1
  r2_2 <- sem0[[i]]$R2$R2_2
  r2_3 <- sem0[[i]]$R2$R2_3
  temp <- data.frame(M1 = r2_1, M2 = r2_2, M3 = r2_3)
  r2_sem <- rbind(r2_sem, temp)  
}

# wilcoxon signed-rank test
## L1
hist(r2_rf$M1 - r2_sem$M1)
wt1 <- wilcox.test(r2_rf$M1, r2_sem$M1, paired = TRUE, alternative = "two.sided")
print(wt1)

## L2
hist(r2_rf$M2 - r2_sem$M2)
wt2 <- wilcox.test(r2_rf$M2, r2_sem$M2, paired = TRUE, alternative = "two.sided")
print(wt2)

## L3
hist(r2_rf$M3 - r2_sem$M3)
wt3 <- wilcox.test(r2_rf$M3, r2_sem$M3, paired = TRUE, alternative = "two.sided")
print(wt3)


# t-test
## L1
tt1 <- t.test(r2_rf$M1, r2_sem$M1, paired = TRUE, alternative = "two.sided")
print(tt1)

## L2
tt2 <- t.test(r2_rf$M2, r2_sem$M2, paired = TRUE, alternative = "two.sided")
print(tt2)

## L3
tt3 <- t.test(r2_rf$M3, r2_sem$M3, paired = TRUE, alternative = "two.sided")
print(tt3)

```

## Combine and plot R2

```{r r2}
# pivot and combine
r2_rf <- r2_rf %>% pivot_longer(cols = c(M1, M2, M3), names_to = "level", values_to = "value")
r2_rf$Model <- "RandomForest"

r2_sem <- r2_sem %>% pivot_longer(cols = c(M1, M2, M3), names_to = "level", values_to = "value")
r2_sem$Model <- "PiecewiseSEM"

r2_all <- rbind(r2_rf, r2_sem)
r2_all <- r2_all %>% group_by(Model, level) %>% 
  summarise(
    est = mean(value),
    se = sd(value) / sqrt(n()),
    t = qt(0.975, df = n() - 1),
    hi = est + t * se,
    lo = est - t * se
  )
r2_all$labels <- if_else(r2_all$level=="M1", "log(Turbidity)",
                         if_else(r2_all$level=="M2", "log(Chl-a)", "Macroalgae"))
r2_all$labels <- factor(r2_all$labels, levels = c("Macroalgae","log(Chl-a)","log(Turbidity)"))

# plot
w <- 0.5
p <- ggplot(r2_all, aes(x = labels, y = est, color = Model)) +
  geom_point(position = position_dodge(width = w)) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = w, position = position_dodge(width = w)) +
  ylim(0,1) +
  coord_flip() +
  theme_minimal() +
  labs(
    y = "R-squared",
    x = ""
  )
ggsave("figures/r2_plot.png", plot = p, width = 5, height = 3, units = "in")
```

# Feature Importance

```{r feats}
library(data.table)

feats_rf <- vector("list", B*(3+15+15))
pdps <- vector("list", B*(3+15+15))
psem_0 <- vector("list", B)

k <- 0L
for (i in seq(B)) {
  # forest
  ## L1
  vs <- names(rf1[[i]]$imp)
  for (j in seq_along(vs)) {
    k <- k + 1L
    feats_rf[[k]] <- data.table(
      Response = "log_turb",
      Model = "forest",
      Variable = vs[j],
      Importance = rf1[[i]]$imp[[vs[j]]]
    )

    pdps[[k]] <- data.table(
      Response = "log_turb",
      Ind.Var = rf1[[i]]$pdp[[j]]$Var,
      Value = rf1[[i]]$pdp[[j]]$PDP[[1]],
      yhat = rf1[[i]]$pdp[[j]]$PDP$yhat
    )
  }
  
  ## L2
  vs <- names(rf2[[i]]$imp)
  for (j in seq_along(vs)) {
    k <- k + 1L
    feats_rf[[k]] <- data.table(
      Response = "log_chla",
      Model = "forest",
      Variable = vs[j],
      Importance = rf2[[i]]$imp[[vs[j]]]
    )

    pdps[[k]] <- data.table(
      Response = "log_chla",
      Ind.Var = rf2[[i]]$pdp[[j]]$Var,
      Value = rf2[[i]]$pdp[[j]]$PDP[[1]],
      yhat = rf2[[i]]$pdp[[j]]$PDP$yhat
    )
  }
  
  ## L3
  vs <- names(rf3[[i]]$imp)
  for (j in seq_along(vs)) {
    k <- k + 1L
    feats_rf[[k]] <- data.table(
      Response = "AMITOTAL",
      Model = "forest",
      Variable = vs[j],
      Importance = rf3[[i]]$imp[[vs[j]]]
    )

    pdps[[k]] <- data.table(
      Response = "AMITOTAL",
      Ind.Var = rf3[[i]]$pdp[[j]]$Var,
      Value = rf3[[i]]$pdp[[j]]$PDP[[1]],
      yhat = rf3[[i]]$pdp[[j]]$PDP$yhat
    )
  }
  
  # pSEM
  psem_0[[i]] <- data.table(
    Response = sem0[[i]]$imp[["Response"]], 
    Model = "pSEM", 
    Variable = sem0[[i]]$imp[["Predictor"]],
    Importance = sem0[[i]]$imp[["Std.Estimate"]]
  )
}

# aggregate
pdps <- rbindlist(pdps)

feats <- rbindlist(c(psem_0, feats_rf))

agg_feats <- feats[
  , .(
    mean = mean(Importance),
    se = sd(Importance) / sqrt(.N),
    n = .N,
    q_lower = quantile(Importance, 0.025),
    q_upper = quantile(Importance, 0.975)
  ),
  by = .(Model, Response, Variable)
][,
  `:=`(
    lower_ci = mean - qt(0.975, df = n - 1) * se,
    upper_ci = mean + qt(0.975, df = n - 1) * se
  )
]

agg_pdps <- pdps[
  , .(
    yhat.mean = mean(yhat),
    lo95 = quantile(yhat, 0.025),
    hi95 = quantile(yhat, 0.975),
    lo50 = quantile(yhat, 0.25),
    hi50 = quantile(yhat, 0.75)
  ),
  by = .(Response, Ind.Var, Value)
]

# write.csv(agg_feats, paste0(getwd(),"/Data/agg_feats.csv"), row.names = FALSE)
# write.csv(agg_pdps, paste0(getwd(),"/Data/agg_pdps.csv"), row.names = FALSE)

```

## Compare features

```{r feats plot}
# sort
sort_feats <- agg_feats
sort_feats <- sort_feats %>% filter(Response != "~~log_chla")

sort_feats$Response <- factor(sort_feats$Response, levels = c("log_turb","log_chla","AMITOTAL"))
sort_feats$Model <- factor(sort_feats$Model, levels = c("forest","pSEM"))

order_df <- sort_feats[sort_feats$Model == "forest" & sort_feats$Response != "AMITOTAL",]
ordered_levels <- order_df$Variable[order(abs(order_df$mean))]
sort_feats$Variable <- factor(sort_feats$Variable, levels = ordered_levels)

# plot
p <- ggplot(sort_feats %>% filter(Response != "~~log_chla"), 
       aes(x = Variable, y = mean, color = Model)) +
  geom_hline(yintercept = 0, color = "black") +
  geom_pointrange(aes(ymin = q_lower, ymax = q_upper)) +
  facet_grid(Response ~ Model, scales = "free") +
  coord_flip() +
  theme_minimal() +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 1)) +
  labs(
    y = "",
    x = ""
  )
ggsave("figures/importance_comp.png", plot = p, width = 7, height = 9, units = "in")
```

## Partial Dependence
https://rpubs.com/vishal1310/QuickIntroductiontoPartialDependencePlots

"If I fix feature X at value x and average predictions over the observed distribution of all other features, what is the predicted response?" It's a marginal effect — a single curve that summarizes the population-average relationship between one feature and the outcome.
PDPs aggregate more naturally since they produce a curve (feature value → prediction) rather than per-observation values. You can average curves across iterations directly if you evaluate them on a common grid

For your specific goals, quantile bands are more informative and appropriate. Here's why:
You're not trying to quantify uncertainty in a population parameter — you're trying to assess whether the feature's learned relationship is stable across different lake subsets. A narrow quantile band means the RF consistently learns the same shape regardless of which lakes it trains on. A wide band means the relationship is sensitive to which lakes are included — ecologically meaningful information that a CI around the mean would obscure.
A CI around the mean would shrink toward zero as you run more iterations, eventually showing an implausibly tight band that suggests false precision — it conflates iteration count with ecological certainty.

Sparse data at that feature value. The most common explanation. If only a few lakes have TP values in a certain range, the RF has few training examples there and can produce unstable, spiky predictions. The rug plot will show this — a peak in a sparse region is almost certainly a data artifact.

A PDP estimates the marginal effect of one feature on the predicted outcome by integrating out (averaging over) all other features. The core idea is to ask: if we could vary feature X across its range while holding the joint distribution of all other features fixed at their observed values, what would the model predict on average?

```{r}
# get top feats
top_turb <- c("log_chla","calc_clay","RVFPCANDECIDUOUS_RIP")
top_chla <- agg_feats %>% filter(Response=="log_chla", Model=="forest") %>% arrange(desc(mean)) %>% head(5) %>% select(Variable) %>% as.list()
top_macro <- agg_feats %>% filter(Response=="AMITOTAL", Model=="forest") %>% arrange(desc(mean)) %>% head(5) %>% select(Variable) %>% as.list()

# filter pdps
turb_pdps <- agg_pdps %>% filter(Response=="log_turb", Ind.Var %in% top_turb)
turb_pdps$Ind.Var <- factor(turb_pdps$Ind.Var, levels = top_turb)

chla_pdps <- agg_pdps %>% filter(Response=="log_chla", Ind.Var %in% top_chla$Variable)
chla_pdps$Ind.Var <- factor(chla_pdps$Ind.Var, levels = top_chla$Variable)

macro_pdps <- agg_pdps %>% filter(Response=="AMITOTAL", Ind.Var %in% top_macro$Variable)
macro_pdps$Ind.Var <- factor(macro_pdps$Ind.Var, levels = top_macro$Variable)

# plots
p <- ggplot(turb_pdps, aes(x=Value, y=yhat.mean)) +
  geom_ribbon(aes(ymin=lo95, ymax=hi95), alpha=0.2, fill="blue") +
  geom_ribbon(aes(ymin=lo50, ymax=hi50), alpha=0.4, fill="green") +
  geom_line(linewidth = 1, color = "black") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~Ind.Var, scales="free_x") +
  theme_minimal() +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 1)) +
  labs(
    x = "",
    y = "Partial Dependence",
    title = "log(Turbidity) PDPs"
  )
ggsave("figures/pdp_turb.png", plot = p, width = 6, height = 3, units = "in")

p <- ggplot(chla_pdps, aes(x=Value, y=yhat.mean)) +
  geom_ribbon(aes(ymin=lo95, ymax=hi95), alpha=0.2, fill="blue") +
  geom_ribbon(aes(ymin=lo50, ymax=hi50), alpha=0.4, fill="green") +
  geom_line(linewidth = 1, color = "black") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~Ind.Var, scales="free_x", nrow = 1) +
  theme_minimal() +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 1)) +
  labs(
    x = "",
    y = "Partial Dependence",
    title = "log(Chlorophyll-a) PDPs"
  )
ggsave("figures/pdp_chla.png", plot = p, width = 6, height = 3, units = "in")

p <- ggplot(macro_pdps, aes(x=Value, y=yhat.mean)) +
  geom_ribbon(aes(ymin=lo95, ymax=hi95), alpha=0.2, fill="blue") +
  geom_ribbon(aes(ymin=lo50, ymax=hi50), alpha=0.4, fill="green") +
  geom_line(linewidth = 1, color = "black") +
  geom_hline(yintercept = 0, color = "red", linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~Ind.Var, scales="free_x", nrow = 1) +
  theme_minimal() +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 1)) +
  labs(
    x = "",
    y = "Partial Dependence",
    title = "Macroalgae Index PDPs"
  )
ggsave("figures/pdp_macro.png", plot = p, width = 6, height = 3, units = "in")
```

## SHAP
https://cran.r-project.org/web/packages/fastshap/vignettes/fastshap.html
Libraries: shapviz, fastshap

"For this specific observation, how much did each feature push the prediction up or down relative to the baseline (average prediction)?" It's a local, additive decomposition of each individual prediction into per-feature contributions.

```{r shap}
rf.full.1 <- ranger(m1, data = all_vars, num.trees = 500, importance = "permutation")
ind.vars.1 <- labels(terms(m1))
ex <- explain(rf.full.1, X = all_vars[,ind.vars.1], pred_wrapper = pred_ranger, nsim = 10, adjust = TRUE, shap_only = TRUE)
shp <- shapviz(ex, X = all_vars[,ind.vars.1])

sv_waterfall(shp, row_id = seq(N))
sv_dependence2D(shp, x = "log_chla", y = "calc_clay")
sv_importance(shp, kind = "both")
sv_dependence(shp, v = c("log_chla", "calc_clay"), color_var = NULL, share_y = TRUE)

shp.int <- data.frame(
  chl = shp$X$log_chla,
  clay = shp$X$calc_clay,
  SHAP = shp$S[,"log_chla"] + shp$S[,"calc_clay"]
)

ggplot(shp.int, aes(x=chl, y=clay, z=SHAP)) +
  stat_summary_hex(fun = mean) +
  scale_fill_gradient2(midpoint = 0)

ggplot(shp.int, aes(x=chl, y=clay, color=SHAP)) +
  geom_point(alpha=0.8, size=3) +
  scale_color_gradient2(midpoint = 0, low = "paleturquoise", high = "seagreen", mid = "turquoise")

ggplot(shp.int, aes(x=chl, y=clay)) +
  stat_density2d_filled(aes(fill = after_stat(level)), alpha=0.8) +
  # geom_point(aes(color=SHAP), size=1) +
  scale_color_gradient2(midpoint = 0)
```

## Full pSEM

```{r full psem}
# pSEM
all_vars_std <- as.data.frame(scale(all_vars, center = TRUE, scale = TRUE))

lm1 <- lm(m1, data = all_vars_std)
lm2 <- lm(m2, data = all_vars_std)
lm3 <- lm(m3, data = all_vars_std)

s0 <- psem(lm1, lm2, lm3, log_chla %~~% AMITOTAL, data = all_vars_std)
sum0 <- summary(s0, .progressBar = FALSE)
imp <- sum0$coefficients

plot(s0)
plot(all_vars_std$log_chla, all_vars_std$log_turb)
```

